### 2010-07-23 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: uniform = vs. \mapsto: choose \mapsto
Desc@ICFP: helplessly stare at whitespaces and parenthesis
Desc@ICFP: remove duplicate definition of DotTo.
Desc@ICFP: this \Eq is a \PropEq
Desc@ICFP: uniform Desc and IDesc codes presentation
Desc@ICFP: life and section 5 in colour^W color.

We define the $\SYMBIDesc$ grammar in Figure~\ref{fig:idesc},
delivering only \emph{strictly positive} families. As well as
indexing our descriptions, we have refactored a little, adopting
a more compositional algebra of codes, where $\Desc$ is
biased towards the right-nested tuples. We now have
\(\DVar{i}\) for recursive `subrequests' at a chosen index \(i\),
with tupling by right-associative
\(\DProd{}{}\) and higher-order branching
by \(\DPi{\!}{\!}\).  Upgrade your old $\Desc$
to a trivially indexed $\IDesc{\Unit}$ as follows!
\[\begin{array}{@{}ll}
\upgrade :\Desc & \To\IDesc{\Unit} \\
\upgrade\:\DUnit & = \DConst{\Unit} \\
\upgrade\:(\DSigma{\V{S}}{\V{D}}) &
   = \DSigma{\V{S}}{\LAM{\V{s}}\upgrade\:(\V{D}\:\V{s})} \\
\upgrade\:(\DIndx{\V{D}}) & =
  \DProd{\DVar{\Void}}{\upgrade\:\V{D}} \\
\upgrade\:(\DHindx{\V{H}}{\V{D}}) & =
  \DProd{(\DPi{\V{H}}{\LAM{\_}\DVar{\Void}})}{\upgrade\:\V{D}} \\
\end{array}\]

To deliver induction for indexed datatypes, we need the `holds everywhere'
machinery. We present $\SYMBAllI$ and $\SYMBallI$ in
Figure~\ref{fig:allI-predicates}, with a twist---where
$\Desc$ admits the $\SYMBall$ construction, $\SYMBIDesc$ is \emph{closed}
under it! The $\SYMBAllI$
operator for a description indexed on \(\V{I}\) is strictly positive in
turn, and has a description indexed on
 some \(\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}}\).
Induction on indexed descriptions is then hardwired thus:
%
\[\stk{
\begin{array}{@{}ll}
\inductionI : & _{\PI{\V{I}}{\Set}}
                   \PITEL{\V{R}}{\V{I} \To \IDesc{\V{I}}}
                   \PI{\V{P}}{(\SIGMA{\V{i}}{\V{I}}{\IMu{\V{I}}{\V{R}}{\V{i}}}) \To \Set} \\
                 & (      \PITEL{\V{i}}{\V{I}} 
                          \PI{\V{xs}}{\idescop{\V{R}\: \V{i}}{\V{I}}{(\SYMBIMu_{\V{I}}{\V{R}})}} \\
                 & \   \idescop{\AllI{}
                                     {(\V{R}\: \V{i})}
                                     {(\SYMBIMu_{\V{I}}{\V{R}})}
                                     {\V{xs}}}
                               {}
                               {\V{P}} \To
                       \V{P}\: \pair{\V{i}}{\Con{\V{xs}}}{}) \To \\
                 & \PITEL{\V{i}}{\V{I}}
                   \PI{\V{x}}{\IMu{\V{I}}{\V{R}}{\V{i}}}
                   \V{P}\: \pair{\V{i}}{\V{x}}{}
\end{array} \\
\inductionI\: \V{R}\: \V{P}\: \V{m}\: \V{i}\: (\Con{\V{xs}}) =  \\
\qquad
    \V{m}\: \V{i}\: \V{xs}\: (\allI{}
                                    {\V{R}\: \V{i}}
                                    {(\SYMBIMu_{\V{I}}{\V{R}})}
                                    {\V{P}}
                                    {(\spl{\LAM{\V{i}}\LAM{\V{xs}} \inductionI\: \V{R}\: \V{P}\: \V{m}})}
                                    {\V{xs}})
}\]
%
The generic catamorphism, $\cataI$, is constructed from $\inductionI$
as before.
\note{do this later?}

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\SYMBAllI : & _{\PI{\V{I}}{\Set}}
              \PITEL{\V{D}}{\IDesc{\V{I}}}
              \PI{\V{X}}{\V{I} \To \Set} \\
            & \idescop{\V{D}}{\V{I}}{\V{X}} \To
              \IDesc{(\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}})}
\end{array} \\
\begin{array}{@{}l@{}l@{\:=\:\:}l}
\AllI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{x}} &
    \DVar{\pair{\V{i}}{\V{x}}{}} \\
\AllI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{k}} &
    \DConst{\Unit} \\
\AllI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\pair{\V{d}}{\V{d'}}{}} &
    \DProd{\AllI{}{\V{D}}{\V{X}}{\V{d}}}{\AllI{}{\V{D'}}{\V{X}}{\V{d'}}} \\
\AllI{\:}{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\pair{\V{s}}{\V{d}}{}} &
    \AllI{}{(\V{D}\: \V{s})}{\V{X}}{\V{d}} \\
\AllI{\:}{(\DPi{\V{S}}{\V{D}})}{& \V{X}}{\V{f}} &
    \DPi{\V{S}}{(\LAM{\V{s}} \AllI{}{(\V{D}\: \V{s})}{\V{X}}{(\V{f}\: \V{s})})}
\end{array}
}
&
%%
\stk{
\begin{array}{@{}ll}
\SYMBallI : & _{\PI{\V{I}}{\Set}}
              \PITEL{\V{D}}{\IDesc{\V{I}}}
              \PITEL{\V{X}}{\V{I} \To \Set} 
              \PI{\V{P}}{(\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}}) \To \Set} \\
            & (\PI{\V{x}}{\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}}} \V{P}\: \V{x}) \To
              \PI{\V{xs}}{\idescop{\V{D}}{\V{I}}{\V{X}}} 
              \idescop{\AllI{}{\V{D}}{\V{X}}{\V{xs}}}{}{\V{P}}
\end{array} \\
\begin{array}{@{}l@{}l@{\:=\:\:}l}
\allI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{P}}{\V{p}}{\V{x}} &
    \V{p}\: \pair{\V{i}}{\V{x}}{} \\
\allI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{P}}{\V{p}}{\V{k}} &
    \void \\
\allI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{d}}{\V{d'}}{}} &
    \pair{\allI{}{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}}
         {\allI{}{\V{D'}}{\V{X}}{\V{P}}{\V{p}}{\V{d'}}}{} \\
\allI{\:}{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{s}}{\V{d}}{}} &
    \allI{}{(\V{D}\: \V{s})}{\V{X}}{\V{P}}{\V{p}}{\V{d}} \\
\allI{\:}{(\DPi{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\V{f}} &
    \LAM{\V{a}}\allI{}{(\V{D}\: \V{a})}{\V{X}}{\V{P}}{\V{p}}{(\V{f}\: \V{a})}
\end{array}
\end{array}
}
\]

@@

\[\stk{\begin{array}{@{}ll}
\IDesc{(\Bhab{\M{I}}{\Set})} &: \Set \\
\DVar{(\Bhab{\M{i}}{\M{I}})} &: \IDesc{\M{I}} \\
\DConst{(\Bhab{\M{A}}{\Set})} &:\IDesc{\M{I}}       \\
\DProd{(\Bhab{\M{D}}{\IDesc{\M{I}}})}{(\Bhab{\M{D}}{\IDesc{\M{I}}})}
  & :\IDesc{\M{I}}       \\
\DSigma{(\Bhab{\M{S}}{\Set})}{(\Bhab{\M{D}}{\M{S}\To\IDesc{\M{I}}})}
& : \IDesc{\M{I}}  \\
\DPi{(\Bhab{\M{S}}{\Set})}{(\Bhab{\M{D}}{\M{S}\To\IDesc{\M{I}}})}
& : \IDesc{\M{I}}  \\
\end{array}\smallskip \\
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\V{X}}@{\:\mapsto\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{&}      &  \V{X}\: \V{i}                                           \\
\idescop{\DConst{\V{K}}}{\V{I}}{&}    &  \V{K}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{&} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{X}}}{\idescop{\V{D'}}{\V{I}}{\V{X}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{&} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{X}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{&}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{X}}            
\end{array}
}
\]

@@

For basic reassurance, we \(\upgrade\:\NatD\):
%
\[\stk{
\upgrade\:\NatD : \IDesc{\Unit} \\
\upgrade\:\NatD = \DSigma{(\EnumT{[ \NatZero\: \SYMBNatSuc ]})}
                     {\red{[} (\DConst{\Unit}) \; 
   (\DProd{\DVar{\Void}}{\DConst{\Unit}}) \red{]}}
}\]
%
Note that trailing \(\Unit\)'s keep our right-nested, \(\void\)-terminated
tuple structure, and with it our elaboration machinery.
We can similarly \(\upgrade\) any inductive type.
Moreover, \(\IDesc{I}\) can now code a bunch of mutually
inductive types, if \(I\) enumerates the
bunch~\cite{paulin:habilitation, yakushev:mutual-def}.

@@

One way to code constrained datatypes is to appeal to a suitable
notion of propositional equality \(\PropEq\) on indices. The
constraints are expressed as `Henry Ford' equations in the datatype.
For vectors:
%
\[\stk{
\VecD : \Set \To \Nat \To \IDesc{\Nat} \\
\VecD\: \V{X}\: \V{i} = \SYMBDSigma\\
\quad
\EnumT{\vtup{r}{\VecNil\\ \SYMBVecCons}}
\; \vtup{r}{
                            (\DConst{(\NatZero\PropEq\V{i})}) \\
 ( \DSigma{\Nat}{\LAM{\V{n}}
   \DProd{\DConst{\V{X}}}
     {\DProd{\DVar{\V{n}}}{\DConst{(\NatSuc{\V{n}}\PropEq\V{i})}})}}
                          }
}\]

@@

}\]
%
The constructors and equations have simply disappeared. A similar
example is $\SYMBFin$ (bounded numbers), specified by:
%
\[
\stk{
\data \SYMBFin : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero      & _{\PI{\V{n}}{\Nat}}\Fin{(\NatSuc{\V{n}})}   \\
    \SYMBFinSuc   & _{\PI{\V{n}}{\Nat}}\Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
\end{array}
}\]
%
In this case, we can eliminate equations but not constructors, since both
$\FinZero$ and $\SYMBFinSuc$ both target $\SYMBNatSuc$:
%
\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}lll}
\FinD\: \NatZero         & \mapsto & \DSigma{\EnumT{\Void}}{\Void} \\
\FinD\: (\NatSuc{\V{n}}) & \mapsto & \DSigma
 {\EnumT{\vtup{r}{\FinZero \\ \SYMBFinSuc }}}
 {\vtup{l}{ (\DConst{\Unit}) \\
            (\DVar{\V{n}})}}
\end{array}
}\]

@@

We use the notation $\toIDesc{\V{D}}$ to denote the indexed
description computed from the tagged indexed description $\V{D}$. Its
expansion is similar to the definition of \(\SYMBtoDesc\) for tagged
descriptions. For brevity, we shall also overload $\IMu{I}{D}$ for
$\IMu{I}{(\toIDesc{D})}$, as the bidirectional presentation let us
resolve the ambiguity.

@@

An expression of this language is either a value, a conditional
expression, an addition of numbers, or a comparison of
numbers. Informally, their type is the following:
%
\[
\begin{array}{l@{\::\:\:}l}
\SYMBEcond           & \forall \Bhab{\V{ty}}{\Ty} . \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
\SYMBEplus           & \Enat \To \Enat \To \Enat                           \\
\SYMBEle             & \Enat \To \Enat \To \Ebool                          \\
\SYMBEval            & \forall \Bhab{\V{ty}}{\Ty} . \Val{\V{ty}} \To \V{ty}
\end{array}
\]
%
The function $\SYMBVal$, used in the definition of $\SYMBEval$, simply
maps a type $ty$ of the object language to the corresponding type in
the host language. Hence, the arguments of $\SYMBEval$ are ensured to
be of the expected type. We assume that $\Nat$ and $\Bool$ represent
natural numbers and booleans in the host language, equipped with an
addition operation $\F{plusHost}$ and a comparison function
$\F{leHost}$. We define $\SYMBVal$ as follows:
%
\[\stk{
\SYMBVal : \Ty \To \Set \\
\begin{array}{@{}l@{\:=\:\:}l}
\Val{\Enat}   & \Nat \\
\Val{\Ebool}  & \Bool
\end{array}
}\]

@@

Having implemented the syntax, we would like to describe its
semantics. To do so, we implement an evaluator. The type of the
evaluator is:
%
\[
\evalH : \PI{\V{ty}}{\Ty} 
         \IMu{\Ty}{\HExprD}{\V{ty}} \To
         \Val{\V{ty}}
\]
%
The type of $\evalH$ is strikingly similar to a
catamorphism. Indeed, implementing a single step of evaluation -- the
algebra -- is sufficient, as $\cataI$ gives, for free, the full
evaluator. The implementation is as follows:
%
\[\stk{
\evalOne : \PI{\V{ty}}{\Ty} \idescop{(\toIDesc{\HExprD})\:\V{ty}}{\Ty}{\SYMBVal}
           \To {\Val{\V{ty}}} \\
\begin{array}{@{}l@{}c@{}l@{\:=\:\:}l}
\evalOne\: & \_\: & \pair{\SYMBEval}{\V{x}}{}                                             & \V{x} \\
\evalOne\: & \_\: & \pair{\SYMBEcond}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
\evalOne\: & \_\: & \pair{\SYMBEcond}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{y} \\
\evalOne\: & \Enat\: & \pair{\SYMBEplus}{\pair{\V{x}}{\V{y}}{}}{}                     & \F{plusHost}\: \V{x}\: \V{y} \\
\evalOne\: & \Ebool\: & \pair{\SYMBEle}{\pair{\V{x}}{\V{y}}{}}{}                      & \F{leHost}\: \V{x}\: \V{y} 
\end{array} \\
\\
\evalH : \PI{\V{ty}}{\Ty} 
           \IMu{\Ty}{\HExprD}{\V{ty}} \To
           \Val{\V{ty}} \\
\evalH\: \V{ty}\: \V{term} = \cataI_{\Ty} \:
                                 (\toIDesc{\HExprD})\: 
                                 \SYMBVal\: 
                                 \evalOne\: 
                                 \V{ty}\: 
                                 \V{term}
}\]

@@

The $\discharge$ function is relative to the required type and a
context of the right type. Its action is to map values to themself,
and variables to their value in context. The pointwise indexed
function type $\PI{\V{i}}{I}{X\: i \To Y\: i}$ is abbreviated  $X
\DotTo Y$. This corresponds to the following function:
%
\[\stk{
\begin{array}{@{}ll}
\discharge : & \PI{\V{G}}{\Ctxt}
               \Env{G} \To 
              (\SUM{\Val{\V{}}}{\Var{\V{G}}{}}) \DotTo
               \IMu{\Ty}{\HExprD}{}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumLeft \V{x})  & 
    \Con{\pair{\SYMBEval}{\V{x}}{}} \\
\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumRight \V{v}) &
    \Con{\pair{\SYMBEval}{\lookup{\V{G}}{\V{g}}{\V{ty}}{\V{v}}}{}}
\end{array}
}\]

>>>>

We define the $\SYMBIDesc$ grammar in Figure~\ref{fig:idesc},
delivering only \emph{strictly positive} families. As well as
indexing our descriptions, we have refactored a little, adopting
a more compositional algebra of codes, where $\Desc$ is
biased towards the right-nested tuples. We now have
\(\DVar{i}\) for recursive `subrequests' at a chosen index \(i\),
with tupling by right-associative
\(\DProd{}{}\) and higher-order branching
by \(\DPi{\!}{\!}\).  Upgrade your old $\Desc$
to a trivially indexed $\IDesc{\Unit}$ as follows!
\[\begin{array}{@{}ll}
\upgrade :\Desc & \To\IDesc{\Unit} \\
\upgrade\:\DUnit & \mapsto \DConst{\Unit} \\
\upgrade\:(\DSigma{\V{S}}{\V{D}}) &
   \mapsto \DSigma{\V{S}}{\LAM{\V{s}}\upgrade\:(\V{D}\:\V{s})} \\
\upgrade\:(\DIndx{\V{D}}) & \mapsto
  \DProd{\DVar{\Void}}{\upgrade\:\V{D}} \\
\upgrade\:(\DHindx{\V{H}}{\V{D}}) & \mapsto
  \DProd{(\DPi{\V{H}}{\LAM{\_}\DVar{\Void}})}{\upgrade\:\V{D}} \\
\end{array}\]

To deliver induction for indexed datatypes, we need the `holds everywhere'
machinery. We present $\SYMBAllI$ and $\SYMBallI$ in
Figure~\ref{fig:allI-predicates}, with a twist---where
$\Desc$ admits the $\SYMBall$ construction, $\SYMBIDesc$ is \emph{closed}
under it! The $\SYMBAllI$
operator for a description indexed on \(\V{I}\) is strictly positive in
turn, and has a description indexed on
 some \(\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}}\).
Induction on indexed descriptions is then hardwired thus:
%
\[\stk{
\begin{array}{@{}ll}
\inductionI : & _{\PI{\V{I}}{\Set}}
                   \PITEL{\V{R}}{\V{I} \To \IDesc{\V{I}}}
                   \PI{\V{P}}{(\SIGMA{\V{i}}{\V{I}}{\IMu{\V{I}}{\V{R}}{\V{i}}}) \To \Set} \\
                 & (      \PITEL{\V{i}}{\V{I}} 
                          \PI{\V{xs}}{\idescop{\V{R}\: \V{i}}{\V{I}}{(\SYMBIMu_{\V{I}}{\V{R}})}} \\
                 & \   \idescop{\AllI{}
                                     {(\V{R}\: \V{i})}
                                     {(\SYMBIMu_{\V{I}}{\V{R}})}
                                     {\V{xs}}}
                               {}
                               {\V{P}} \To
                       \V{P}\: \pair{\V{i}}{\Con{\V{xs}}}{}) \To \\
                 & \PITEL{\V{i}}{\V{I}}
                   \PI{\V{x}}{\IMu{\V{I}}{\V{R}}{\V{i}}}
                   \V{P}\: \pair{\V{i}}{\V{x}}{}
\end{array} \\
\inductionI\: \V{R}\: \V{P}\: \V{m}\: \V{i}\: (\Con{\V{xs}}) \mapsto 
    \V{m}\: \V{i}\: \V{xs}\: (\allI{}
                                    {\V{R}\: \V{i}}
                                    {(\SYMBIMu_{\V{I}}{\V{R}})}
                                    {\V{P}}
                                    {(\spl{\LAM{\V{i}}\LAM{\V{xs}} \inductionI\: \V{R}\: \V{P}\: \V{m}})}
                                    {\V{xs}})
}\]
%
The generic catamorphism, $\cataI$, is constructed from $\inductionI$
as before.
\note{do this later?}

@@

\[
\begin{array}{ll}
%%
\stk{
\begin{array}{@{}ll}
\SYMBAllI : & _{\PI{\V{I}}{\Set}}
              \PITEL{\V{D}}{\IDesc{\V{I}}}
              \PI{\V{X}}{\V{I} \To \Set} \\
            & \idescop{\V{D}}{\V{I}}{\V{X}} \To
              \IDesc{(\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}})}
\end{array} \\
\begin{array}{@{}l@{}l@{\:\mapsto\:\:}l}
\AllI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{x}} &
    \DVar{\pair{\V{i}}{\V{x}}{}} \\
\AllI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{k}} &
    \DConst{\Unit} \\
\AllI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\pair{\V{d}}{\V{d'}}{}} &
    \DProd{\AllI{}{\V{D}}{\V{X}}{\V{d}}}{\AllI{}{\V{D'}}{\V{X}}{\V{d'}}} \\
\AllI{\:}{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\pair{\V{s}}{\V{d}}{}} &
    \AllI{}{(\V{D}\: \V{s})}{\V{X}}{\V{d}} \\
\AllI{\:}{(\DPi{\V{S}}{\V{D}})}{& \V{X}}{\V{f}} &
    \DPi{\V{S}}{\LAM{\V{s}} \AllI{}{(\V{D}\: \V{s})}{\V{X}}{(\V{f}\: \V{s})}}
\end{array}
}
&
%%
\stk{
\begin{array}{@{}ll}
\SYMBallI : & _{\PI{\V{I}}{\Set}}
              \PITEL{\V{D}}{\IDesc{\V{I}}}
              \PITEL{\V{X}}{\V{I} \To \Set} 
              \PI{\V{P}}{(\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}}) \To \Set} \\
            & (\PI{\V{x}}{\SIGMA{\V{i}}{\V{I}}{\V{X}\: \V{i}}} \V{P}\: \V{x}) \To
              \PI{\V{xs}}{\idescop{\V{D}}{\V{I}}{\V{X}}} 
              \idescop{\AllI{}{\V{D}}{\V{X}}{\V{xs}}}{}{\V{P}}
\end{array} \\
\begin{array}{@{}l@{}l@{\:\mapsto\:\:}l}
\allI{\:}{(\DVar{\V{i}})}{& \V{X}}{\V{P}}{\V{p}}{\V{x}} &
    \V{p}\: \pair{\V{i}}{\V{x}}{} \\
\allI{\:}{(\DConst{\V{K}})}{& \V{X}}{\V{P}}{\V{p}}{\V{k}} &
    \void \\
\allI{\:}{(\DProd{\V{D}}{\V{D'}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{d}}{\V{d'}}{}} &
    \pair{\allI{}{\V{D}}{\V{X}}{\V{P}}{\V{p}}{\V{d}}}
         {\allI{}{\V{D'}}{\V{X}}{\V{P}}{\V{p}}{\V{d'}}}{} \\
\allI{\:}{(\DSigma{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\pair{\V{s}}{\V{d}}{}} &
    \allI{}{(\V{D}\: \V{s})}{\V{X}}{\V{P}}{\V{p}}{\V{d}} \\
\allI{\:}{(\DPi{\V{S}}{\V{D}})}{& \V{X}}{\V{P}}{\V{p}}{\V{f}} &
    \LAM{\V{a}}\allI{}{(\V{D}\: \V{a})}{\V{X}}{\V{P}}{\V{p}}{(\V{f}\: \V{a})}
\end{array}
\end{array}
}
\]

@@

\[\stk{\begin{array}{@{}ll}
\IDesc{(\Bhab{\V{I}}{\Set})} &: \Set \\
\DVar{(\Bhab{\V{i}}{\V{I}})} &: \IDesc{\V{I}} \\
\DConst{(\Bhab{\V{A}}{\Set})} &:\IDesc{\V{I}}       \\
\DProd{(\Bhab{\V{D}}{\IDesc{\V{I}}})}{(\Bhab{\V{D}}{\IDesc{\V{I}}})}
  & :\IDesc{\V{I}}       \\
\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S}\To\IDesc{\V{I}}})}
& : \IDesc{\V{I}}  \\
\DPi{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S}\To\IDesc{\V{I}}})}
& : \IDesc{\V{I}}  \\
\end{array}\smallskip \\
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\V{X}}@{\:\mapsto\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{&}      &  \V{X}\: \V{i}                                           \\
\idescop{\DConst{\V{K}}}{\V{I}}{&}    &  \V{K}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{&} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{X}}}{\idescop{\V{D'}}{\V{I}}{\V{X}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{&} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{X}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{&}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{X}}            
\end{array}
}
\]

@@

For basic reassurance, we \(\upgrade\:\NatD\):
%
\[\stk{
\upgrade\:\NatD : \IDesc{\Unit} \\
\upgrade\:\NatD \mapsto \DSigma{(\EnumT{\sqr{\NatZero\: \SYMBNatSuc}})}
                     {\sqr{(\DConst{\Unit}) \; 
                           (\DProd{\DVar{\Void}}{\DConst{\Unit}})}}
}\]
%
Note that trailing \(\Unit\)'s keep our right-nested, \(\void\)-terminated
tuple structure, and with it our elaboration machinery.
We can similarly \(\upgrade\) any inductive type.
Moreover, \(\IDesc{I}\) can now code a bunch of mutually
inductive types, if \(I\) enumerates the
bunch~\cite{paulin:habilitation, yakushev:mutual-def}.

@@

One way to code constrained datatypes is to appeal to a suitable
notion of propositional equality \(\PropEq\) on indices. The
constraints are expressed as `Henry Ford' equations in the datatype.
For vectors:
%
\[\stk{
\VecD : \Set \To \Nat \To \IDesc{\Nat} \\
\VecD\: \V{X}\: \V{i} \mapsto \SYMBDSigma\\
\quad
\EnumT{\vtup{r}{\VecNil\\ \SYMBVecCons}}
\; \vtup{r}{
                            (\DConst{(\NatZero\PropEq\V{i})}) \\
 ( \DSigma{\Nat}{\LAM{\V{n}}
   \DProd{\DConst{\V{X}}}
     {\DProd{\DVar{\V{n}}}{\DConst{(\NatSuc{\V{n}}\PropEq\V{i})}})}}
                          }
}\]

@@

}\]
%
The constructors and equations have simply disappeared. A similar
example is $\SYMBFin$ (bounded numbers), specified by:
%
\[
\stk{
\data \SYMBFin : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero      & _{\PI{\V{n}}{\Nat}}\Fin{(\NatSuc{\V{n}})}   \\
    \SYMBFinSuc   & _{\PI{\V{n}}{\Nat}}\Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
\end{array}
}\]
%
In this case, we can eliminate equations but not constructors, since both
$\FinZero$ and $\SYMBFinSuc$ both target $\SYMBNatSuc$:
%
\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}lll}
\FinD\: \NatZero         & \mapsto & \DSigma{\EnumT{\Void}}{\Void} \\
\FinD\: (\NatSuc{\V{n}}) & \mapsto & \DSigma{\EnumT{\sqr{\FinZero\: \SYMBFinSuc}}}
                                            {\sqr{(\DConst{\Unit})\: (\DVar{\V{n}})}}
\end{array}
}\]

@@

We use the notation $\toIDesc{\V{D}}$ to denote the indexed
description computed from the tagged indexed description $\V{D}$. Its
expansion is similar to the definition of \(\SYMBtoDesc\) for tagged
descriptions. For brevity, we shall also overload $\IMu{\V{I}}{\V{D}}$ for
$\IMu{\V{I}}{(\toIDesc{\V{D}})}$, as the bidirectional presentation let us
resolve the ambiguity.

@@

An expression of this language is either a value, a conditional
expression, an addition of numbers, or a comparison of
numbers. Informally, their type is the following:
%
\[
\begin{array}{l@{\::\:\:}l}
\SYMBEcond           & \forall \Bhab{\V{ty}}{\Ty} . \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
\SYMBEplus           & \Enat \To \Enat \To \Enat                           \\
\SYMBEle             & \Enat \To \Enat \To \Ebool                          \\
\SYMBEval            & \forall \Bhab{\V{ty}}{\Ty} . \Val{\V{ty}} \To \V{ty}
\end{array}
\]
%
The function $\SYMBVal$, used in the definition of $\SYMBEval$, simply
maps a type $ty$ of the object language to the corresponding type in
the host language. Hence, the arguments of $\SYMBEval$ are ensured to
be of the expected type. We assume that $\Nat$ and $\Bool$ represent
natural numbers and booleans in the host language, equipped with an
addition operation $\F{plusHost}$ and a comparison function
$\F{leHost}$. We define $\SYMBVal$ as follows:
%
\[\stk{
\SYMBVal : \Ty \To \Set \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\Val{\Enat}   & \Nat \\
\Val{\Ebool}  & \Bool
\end{array}
}\]

@@

Having implemented the syntax, we would like to describe its
semantics. To do so, we implement an evaluator. The type of the
evaluator is:
%
\[
\evalH : \PI{\V{ty}}{\Ty} 
         \IMu{\Ty}{\HExprD}{\V{ty}} \To
         \Val{\V{ty}}
\]
%
The type of $\evalH$ is strikingly similar to a
catamorphism. Indeed, implementing a single step of evaluation -- the
algebra -- is sufficient, as $\cataI$ gives, for free, the full
evaluator. The implementation is as follows:
%
\[\stk{
\evalOne : \PI{\V{ty}}{\Ty} \idescop{(\toIDesc{\HExprD})\:\V{ty}}{\Ty}{\SYMBVal}
           \To {\Val{\V{ty}}} \\
\begin{array}{@{}l@{}c@{}l@{\:\mapsto\:\:}l}
\evalOne\: & \_\: & \pair{\SYMBEval}{\V{x}}{}                                             & \V{x} \\
\evalOne\: & \_\: & \pair{\SYMBEcond}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
\evalOne\: & \_\: & \pair{\SYMBEcond}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{y} \\
\evalOne\: & \Enat\: & \pair{\SYMBEplus}{\pair{\V{x}}{\V{y}}{}}{}                     & \F{plusHost}\: \V{x}\: \V{y} \\
\evalOne\: & \Ebool\: & \pair{\SYMBEle}{\pair{\V{x}}{\V{y}}{}}{}                      & \F{leHost}\: \V{x}\: \V{y} 
\end{array} \\
\\
\evalH : \PI{\V{ty}}{\Ty} 
           \IMu{\Ty}{\HExprD}{\V{ty}} \To
           \Val{\V{ty}} \\
\evalH\: \V{ty}\: \V{term} \mapsto \cataI_{\Ty} \:
                                 (\toIDesc{\HExprD})\: 
                                 \SYMBVal\: 
                                 \evalOne\: 
                                 \V{ty}\: 
                                 \V{term}
}\]

@@

The $\discharge$ function is relative to the required type and a
context of the right type. Its action is to map values to themself,
and variables to their value in context. This corresponds to the
following function:
%
\[\stk{
\begin{array}{@{}ll}
\discharge : & \PI{\V{G}}{\Ctxt}
               \Env{\V{G}} \To 
              (\SUM{\Val{\V{}}}{\Var{\V{G}}{}}) \DotTo
               \IMu{\Ty}{\HExprD}{}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumLeft \V{x})  & 
    \Con{\pair{\SYMBEval}{\V{x}}{}} \\
\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumRight \V{v}) &
    \Con{\pair{\SYMBEval}{\lookup{\V{G}}{\V{g}}{\V{ty}}{\V{v}}}{}}
\end{array}
}\]
